Innovation and elaboration on the avian tree of life - supplementary materials

### 2023-06-01 Manuscript/main.tex
Done! Pre-check

@@

>>>>

\section{Materials and Methods}

\textbf{Beak shape data.} We used the dataset from \cite{cooney2017mega,hughes2022global,chira2020signature} which consists of four landmarks and three curves of 25 semi-landmarks each, taken from 3D scans of the beaks of 8748 bird species.
The landmark data were aligned with Procrustes superimposition using the R package \texttt{Morpho} \cite{Rcore,Morpho} and we removed the size component and controlled for symmetry (see \cite{cooney2017mega,hughes2022global,chira2020signature} for detailed descriptions of data collection and processing).
We ordinated the superimposed landmarks using a principal components analysis (PCA) and selected the eight first dimensions of the resulting PCA.
These eight dimensions contained at least 95\% of variance in each super-order, and 98.7\% of the variance in the whole dataset.
We used a threshold of $>95$\% of variance explained for each super-order because although the overall distribution of the variance on each PC axis in the whole trait-space is decreasing (i.e. $\sigma$ PC1 $<$ $\sigma$ PC2 $<$ $\sigma$ PC3), this is not always true for each super-order.
For example, in the Mirandornithes, PC2, then PC4, and then PC1 contained the most variance (Fig.
\ref{Fig:ellipses} and S\ref{Fig:axes_variance}).
This protocol for selecting the number of PC axes ensures that the resulting trait-space contains enough variance for each super-order (see Fig. \ref{Fig:axes_variance}).
This procedure resulted in an $8748 \times 8$ matrix that contains at least 95\% of the variance in all observed living bird beaks (hereafter, the shapespace). %@L_shapespace_description

\textbf{Phylogenetic data.} We used a random subsample of 1000 trees from the avian tree distribution of \cite{jetz2012global} for the analyses and one randomly selected tree for the figures (Fig. \ref{Fig:phylogeny}).
We pruned these trees to contain only the 8748 species present in our beak shapespace.

\textbf{Phylogenetic GLMM analysis.} We ran a Bayesian phylogenetic generalised linear mixed model (pGLMM) using the \texttt{MCMCglmm} and \texttt{ape} R packages \cite{MCMCglmm, ape} with the shapespace data as a multidimensional response variable, the error in the dataset as residuals, and each clade's phylogeny as a nested random term, i.e., a model of:

\begin{equation}
\text{shapespace} \mathtt{\sim} \text{residuals(shapespace)} + \text{random(phylogeny)}
\end{equation}

This model allows us to estimate an evolutionary rate matrix, here defined as a matrix containing the rate of among-species divergence in species traits that explicitly incorporates phylogeny (\textbf{R} matrix of \cite{Houle2017} and \textbf{B} matrix of \cite{ Machado2020})) for each clade and the phylogeny as a whole, and one for the residuals in the shapespace itself \cite{robinson2013quantifying}.
These evolutionary rate matrices were then used in a base projection-rejection analysis to measure the elaboration and innovation scores (see below).
We ran two separate nested models.
The first model used one random term for the class-wide phylogeny and 35 nested random terms, one for each super-order and order containing more than 15 species in our dataset (Fig.
\ref{Fig:ellipses}).
This resulted in a multidimensional GLMM with a $8748 \times 8$ response variables, one class-wide residual term and 36 random terms.
The second model was fitted to a subset of the dataset containing only the 5229 Passeriformes species with 30 nested random terms, one for each sub-order and families containing more than 15 species in our dataset (Fig.
\ref{fig_ellipses_passeriformes}.
This resulted in a multidimensional GLMM with a $5229 \times 8$ response variables, one order-wide residual term and 30 random terms.
To account for phylogenetic uncertainty we ran the model using different tree topologies from the tree distribution in \cite{jetz2012global}.
Because of the very large size of our model and dataset, however, running the full model on multiple trees was computationally unfeasible (it would require 2 CPU years and 4TB of RAM for each of the models, i.e. the class-wide phylogeny and the Passeriformes models).
Instead, we developed and implemented a ''mini-chains'' MCMCglmm method using the R package \texttt{mcmcmglmmm} \cite{mcmcmcglmmm}.

\textbf{mcmcmcglmmm.} This method runs relatively short Monte-Carlo Markov chains (MCMC) across a large number of different phylogenetic trees and concatenates the resulting short posteriors into one larger posterior distribution that contains the phylogenetic uncertainty information.
We performed the method using the following protocol (see Fig. \ref{Fig:mcmcmcglmm}).
First we ran three independent MCMCglmm chains, hereafter the parameterisation chains, for 50,000 generations with the trait data and the model described above on the consensus tree from the tree distribution, along with flat priors with a belief parameter of 2\%, i.e. with a very low weight on the priors.
Next we extracted the following parameters from the parameterisation chains: a) the conservative burnin average time across the three chains, defined as the highest number of iterations required to first reach the median posterior likelihood value multiplied by 1.1; and b) the mini-chains priors, defined as the median posterior model values from the parameterisation chains with a belief parameter of 5\%.
We then ran 400 mini chains in parallel with the estimated burnin time and priors to run 10 samples past the burnin.
This resulted in 10 exploitable posterior samples for each tree.
We used 400 trees because that was the number of models required to reach an effective sample size (ESS) of at least 200 for all the estimated parameters (see Fig. \ref{Fig:model_ess_all_birds} and S\ref{Fig:model_ess_passeriformes}).
Using this approach, we reduced the computational time to around 45 CPU hours and 9GB of RAM per mini-chain - a 400 fold improvement.
The total analysis took around three CPU years using the shARC cluster from the University of Sheffield.
Code to reproduce the procedure is available in the \href{https://raw.rawgit.net/TGuillerme/mcmcmcglmmm/main/inst/MCMCglmm_mini_chains.html}{\texttt{mcmcmcglmmm} vignette} \cite{mcmcmcglmmm}.

\textbf{Elaboration/innovation scores using projection/rejection analyses.} We used the distributions of the 4000 posterior evolutionary rate matrices to run the projection/rejection analyses to obtain elaboration and innovation scores for bird beaks.
We used linear algebra to interpret elaboration and innovation \textit{sensu} \cite{endler2005animal} by using the major axis of the evolutionary rate matrices (referred to throughout the text as the phylogenetic major axis of beak variation) as the ''line of least resistance'', or, specifically here, the line of elaboration for each random term (corresponding to the elaboration axis for each clade).
We projected either species or another phylogenetic major axis of beak variation onto that clade.
We can then interpret where a species or a clade falls on the phylogenetic major axis of beak variation of reference as its $elaboration_{species}$ score, and how far a species is from that phylogenetic major axis of beak variation as its $innovation_{species}$ score.
We calculated the elaboration and innovation scores using two main approaches: 1) by clades (i.e.
to measure the $elaboration_{clade}$/$innovation_{clade}$ at the megaevolutionary scale)where we calculated the projections of the phylogenetic major axis of beak variation of each clade onto the class-wide phylogenetic major axis of beak variation (Fig. \ref{Fig:ellipses}); and 2) by species (i.e.
to measure the $elaboration_{species}$/$innovation_{species}$ at the macroevolutionary scale) where we calculated the projections of each species onto a) the class-wide phylogenetic major axis of beak variation and b) the phylogenetic major axis of beak variation of their parent clade (Fig. \ref{Fig:phylogeny}).
To make the results easier to interpret, we centered and scaled elaboration and innovation scores onto the center of the phylogenetic major axis of beak variation of interest and made the elaboration values absolute.
In this way, we can interpret an elaboration or innovation score within the 0-1 range to be non exceptional, i.e within the 95\% confidence interval of the evolutionary rate matrix.
The full mathematical procedure is described in detail in the supplementary materials (section \ref{supp_projection} - note that the procedure described in the section \ref{supp_projection} is applied to each individual evolutionary rate matrix for each term in the model, e.g. for each 4000 posterior variance matrices individually for each $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$ or $innovation_{clade}$ score %@L_individual matrices).
Code to reproduce the analyses is available at \href{https://raw.rawgit.net/TGuillerme/dispRity/master/inst/vignettes/Projection_analysis.html}{this \texttt{dispRity} vignette}.

\textbf{Clade orthogonality.} We measured the amount of orthogonality of each random term (i.e.
clade) in our models compared to their parent clade and their parent's parent clade.
For example for Columbiformes, we measured the amount of orthogonality in the evolutionary ratematrices of (i) the Columbimorphae random terms against the phylogeny random terms, (ii) the Columbiformes random terms against the Columbimorphae random terms (parent) and (iii) the Columbiformes random terms the whole phylogeny random terms (parent's parent).
To measure the amount of orthogonality we calculated the angles between the phylogenetic major axis of beak variation of the focal clade and their parent clade for each posterior evolutionary rate matrix and within each clade, by randomly comparing phylogenetic major axes of beak variation within a clade.
We converted each angle measurement into a degree of orthogonality metric where 0 corresponds to flat angles ($0^\circ$ or  $180^\circ$) and 1 to right angles ($90^\circ$ or $270^\circ$).
We then measured the posterior probability of the angles within a clade being different to the angles among clades.
These results are available in Fig. \ref{Fig:orthogonality}.
Note that we did not use the random skewers method here \cite{cheverud2007research} because we were interested in measuring the amount of orthogonality between pairs of evolutionary rate matrices rather than simply measuring whether they were different (but see supplementary materials for random skewers correlation results \ref{Fig:random_skewers}). %@L_noskewers

\textbf{Phylogenetic generalised least squares analysis.} We ran phylogenetic generalised least squares models (PGLS) to test the effects of elaboration and innovation on beak shape divergence.
Beak shape divergence was calculated as the Euclidean distance of each species beak shape from the centroid of beak shape of all birds.
We then fitted a series of nested models with distance as the response and $elaboration_{species}$, $innovation_{species}$, $elaboration_{clade}$, and $innovation_{clade}$ as predictors.
The most complex model included all two-way interaction terms:

\begin{equation}
\text{distance} \mathtt{\sim} (elaboration_{species} + innovation_{species}) \times (elaboration_{clade} + innovation_{clade})
\end{equation}

We then sequentially removed terms to examine effects on model fit.
We initially estimated Pagel's $\lambda$ \cite{pagel1997inferring} for all models.
However, model comparison based on AIC scores can be misleading when estimating $\lambda$ (i.e. model differences can arise because of different $\lambda$ fits, rather than due to effects of the parameters of interest).
We therefore repeated all models a further two times, once fixing $\lambda$ to 1 (equivalent to assuming Brownian motion model of evolution) and once fixing $\lambda$ to the lowest value estimated from any of the models.
The results were qualitatively similar regardless of $\lambda$ and we report only models based on pure Brownian motion in the main text.
In addition, we estimated Pagel’s $\lambda$ for $elaboration_{species}$ and $innovation_{species}$ as univariate traits.
All models were fitted using the phylolm library using a single phylogeny \cite{phylom}.

@@

\textbf{Acknowledgements.}
We thank Masahito Tusboi and three other anonymous reviewers for their suggestions that greatly helped improving the manuscript.
